---
title: "Easy Webscraping with rvest Workshop"
author: "Konner Macias"
date: "December 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Our first steps:
- Open RStudio
- Set your working directory
- Create a .R (script) or .Rmd (Markdown/Notebook if you want to see your visualizations)

To Create a code snippet in a Markdown, press the insert button above^ or press Ctrl+Alt+I
```{r}
a <- 5
print(a)
```
To run snippet, click the green arrow button to run that chunk.

# Scraping Data
There are a lot of different methods out there, some are pretty nasty because you are working with CSS and HTML code.  
- Today we will go over one method, using a R package called rvest and a google chrome add-on

## Steps
If it's your first time with Rstudio and you don't have any added on packages yet, follow these steps:  
1. Click on "Packages"->"Install"->type in "xml2" and press "install"  
2. Repeat for "httr" and "rvest"  
3. Now let's get an extension to help us scrape web data. Go to: https://chrome.google.com/webstore/detail/selectorgadget/mhjhnkcfbdhnjickkkdbjoemdmbfginb?hl=en
  

Once you see the icon, you're ready to go!  


# Our Simple Sample Project: The Growth of James Harden
## Steps
1. Go to basketball-reference.com and type in "James Harden", go to his page. Note it is possible to get the csv for the data we want, but we are practicing scraping here!
2. Click on the Selector Gadget you just added
3. Click on the table of his career Per Game data
4. Click empty cells to include and click on the yellow blocks that you don't want to include.
5. Keep clicking until you have the entire data frame that you want highlighted!  

V Save this Recipe! V
```{r}
library(xml2)
library(rvest)
james <- read_html("https://www.basketball-reference.com/players/h/hardeja01.html") %>%
  html_nodes("#per_game .full_table .left , #per_game .poptip , #per_game .full_table .center , #per_game .full_table .right , #per_game_clone .full_table .left , #per_game_clone .center") %>%
  html_text()
james
```
The only thing you will change in the future is what the parameter is for read_html and html_nodes.

# Clean and Wrange Data
1. We want to create a clean data frame
```{r}
# Get column names
col_names <- james[1:30]
# Create empty data frame
df <- data.frame(matrix(ncol = 30, nrow = 0))
# Put column names in data frame
colnames(df) <- col_names
# Now we need to add everything else, we know there's 30 columns and 9 rows per observation
# A for loop will do this for us
for (i in 1:9){
  df[i,] <- james[(30*i + 1):(30*i+30)]
}

df
```

```{r}

# The class or type of each column is 'character'
# In order to deal with numbers we must convert to numeric
# Our second, and 6-30 columns must be converted
df[,c(2,6:30)] <- sapply(df[,c(2,6:30)], as.numeric)
df
```
# Awesome we have a clean data frame!
Now we can create a quick bar chart of his progress in Pts/Game
```{r}
# Plot his Pts/Game
barplot(df$PTS, names.arg = df$Season, main = "Harden's PPG", xlab = "Season")
```
## Work on Spicing up Visuals!
Download packages like ggplot2 for better visuals!  
# Let's create look at look at everything together
```{r}
# too see multiple plots in one, we use this
par(mfrow=c(2,2))
# Try to memorize this!
# Let's copy what we just made as one of our plots
barplot(df$PTS, names.arg = df$Season, main = "Harden's PPG", xlab = "Season")
# Let's look at his 3 Pt attempts, 3Pt%, and   over time
barplot(df$'3PA', names.arg = df$Season, main = "Harden's 3PA/G", xlab = "Season")
barplot(df$'3P%', names.arg = df$Season, main = "Harden's 3P%/G", xlab = "Season")
barplot(df$AST, names.arg = df$Season, main = "Harden's Ast/G", xlab = "Season")
```

# OKC vs Houston
1. Install "fmsb" package. We're gonna create a Spider chart
```{r}
library(fmsb)
# Stats we want to look at: FG%, 3PA, PTS, AST, STL
labs = c("FG%", "3PA", "PPG", "AST", "STL")
# let's select data only for those stats, so we look at the column numbers
# create two different data frames
red.df <- df[,c(11,13,30,25,26)]
okc <- red.df[1:3,]
houston <- red.df[4:9,]
# get their team colors for visualization
colors <- c("#007AC1","#CE1141")
# To create a radar chart, we need the max and min from each column
maxs <- unname(apply(okc, 2, function(x) max(x, na.rm = TRUE)))
# get the means, max, and min and combine in one df
data = data.frame(matrix(ncol=5,nrow=0))
colnames(data) <- labs
data[1,] = unname(apply(red.df, 2, function(x) max(x, na.rm = TRUE))) # max
data[2,] = unname(apply(red.df, 2, function(x) min(x, na.rm = TRUE))) # min
data[3,] = unname(apply(okc, 2, function(x) mean(x, na.rm = TRUE))) # mean of OKC
data[4,] = unname(apply(houston, 2, function(x) mean(x, na.rm = TRUE))) # mean with OKC

# change last row names
rownames(data)[3:4] <- c("OKC","Houston")

radarchart(data, title = "James Harden", pcol = colors)
legend(-2,0, legend=c("OKC","Houston"),pch=c(15,16), col=colors, lty=c(1,2))
```




